---
title: "Analyzing Grocery Pricing Trends of Becel Products from Vendors in Toronto"
subtitle: "While Voila, Metro, and No Frills may position their prices for Becel products within a similar range, their respective Becel pricing strategies do not appear to be correlated with one another"
author: 
  - Julia Lee
  - Vandan Patel
  - Haowei Fan
  - Tianning He
  - Max LI
  - Shuangyuan Yang
date: today
thanks: "Code and data are available at: https://github.com/JuliaJLee/grocery_vendor_analysis"
date-format: long
abstract: "This paper analyzes grocery product pricing data from three Toronto vendors (Voila, Metro, and No Frills) using data collected through screen-scraping from Project Hammer. Prices were standardized to 100g units to ensure comparability across vendors. The findings show that there is no correlation between the vendors and the Becel products that are sold. However, a slight pattern is seen that most customers buy products that are in between the range $1.10 and $1.30 per 100g. "
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false
#| echo: false
#### Workspace setup ####
library(tidyverse)
library(GGally)
data <- read_csv(here::here("data/data.csv"))
```

```{r}
#| warning: false
#| message: false
#| echo: false
library(stringr)
data$price_numeric <- as.numeric(str_replace_all(data$price, "\\$|/100g", ""))
```

# Introduction

In this analysis, we explore grocery product pricing data from three major Toronto retailers: Voila, Metro, and NoFrills. Using a dataset collected through screen-scraping from Project Hammer, we aim to understand differences in product pricing and distribution patterns across these vendors. The data includes pricing information for various products, which we standardized to 100g units to ensure comparability between stores. By performing data cleaning and visualization, this analysis seeks to compare price distributions across vendors and identify any notable trends.

This present analysis first segments the data by vendor, creating subsets for Voila, Metro, and No Frills. For each subset, we visualize the distribution of product prices at each store. These visualizations provide insights into the pricing strategies of each vendor, highlighting the frequency of different price ranges and enabling a side-by-side comparison of pricing trends across these retailers.

Additionally, the analysis examines whether any relationships exist between the price distributions of the three vendors. This enables a direct comparison of prices for similar products across vendors, facilitating a clearer understanding of how pricing might vary between Voila, Metro, and No Frills. Although no correlation was found specifically for Becel products across vendors, a slight pattern emerged, showing that most customers tend to purchase products priced between $1.10 and $1.30 per 100g.

The remainder of this report details the data used in the analysis (@sec-data), the results of the visualizations (@sec-results), and the conclusions drawn from these comparisons (@sec-discussion), highlighting pricing patterns and potential pricing disparities among these vendors.


# Data {#sec-data}

The preparation and manipulation of the original data is conducted using SQL [@SQL]. The steps to conduct the present analysis on the prepared data are carried out using the statistical programming language, R [@citeR].

The data for this analysis was obtained provided by Project Hammer [@data], and it contains data about product details (e.g. product names, brands, current prices, and old prices, and units for products) for 8 different grocery vendors (Voila, T&T, Loblaws, No Frills, Metro, Galleria, Walmart Canada, and Save-On-Foods). The data is collected through the process of screen-scraping website UIs from vendors within a neighbourhood in Toronto. Data from Project Hammer is available from February 28, 2024, and as this data is collected by extracting information from grocers' websites, it is possible to have missing data, which is more thoroughly addressed in @sec-discussion. Through the method of screen scraping, the original data aims to measure and reflect grocery vendor decisions and trends in product pricing. 

For this analysis, three vendors - Voila, Metro, and No Frills - are considered along with their respective pricing for products that are 100g. The "Price" variable for each vendor in this analysis reflects the current price of their products at the time in which the data was downloaded. The variable, "Units", represents the size of products measured using weight metrics (i.e. grams or kilograms) and counts (i.e. the number of items within a given product). In this case, a weight of 100g is selected to ensure consistency in units across products and vendors. It is also important to highlight that this analysis only considers products from the brand, Becel, across the three vendors.

\newpage

# Results {#sec-results}

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: fig-scatter
#| fig-cap: The Relationships between Becel Product Prices from fVoila, Metro, and No Frills

voila_data <- subset(data, vendor == "Voila")
Metro_data <- subset(data, vendor == "Metro")
NoFrills_data <- subset(data, vendor == "NoFrills")

c1 <- voila_data$price_numeric
c2 <- Metro_data$price_numeric
c3 <- NoFrills_data$price_numeric

min_length <- min(length(c1), length(c2), length(c3))
voila_price <- c1[sample(length(c1), min_length)]
metro_price <- c2[sample(length(c2), min_length)]
nofrills_price <- c3[sample(length(c3), min_length)]
data2 <- data.frame(voila_price, metro_price, nofrills_price)

pairs(data2)
```

@fig-scatter shows a pairwise scatter plot matrix of the price distribution of Voila, Metro, and No Frills. It can be seen from the figure that the prices of Voila and Metro are mainly concentrated between 1.0 and 1.3, while the price of No Frills is bimodal, with peaks at 0.9 and 1.2 respectively. Overall, there is a certain overlap in the pricing strategies of the three brands (in the range of 1.0 to 1.3). The prices of Voila and Metro are more concentrated, which may indicate that their pricing is more uniform and their target markets may be similar. However, No Frills has some differences from them. The bimodal distribution of No Frills' prices reflects that it has formulated diversified pricing strategies for different consumer groups, which may attract low-price consumers and cover part of the mid-to-high-price market. This difference indicates that the three may have different competitive strategies in terms of market positioning and target consumers.

\newpage

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: fig-voila
#| fig-cap: Distribution of the Number of Becel Products Sold Across Different Prices from Voila

voila_data <- subset(data, vendor == "Voila")


plot_voila <- ggplot(voila_data, aes(x = price_numeric)) +
  geom_histogram(binwidth = 0.1, fill = "blue", color = "black", alpha = 0.7) +
  labs(title = "Voila",
       x = "Price/100g",
       y = "Number of Products Bought") +
  theme_minimal()


print(plot_voila)
```
@fig-voila illustrates the distribution of Voila’s price per 100g. The horizontal axis represents the price per 100g, while the vertical axis denotes the frequency of occurrence. The distribution is right-skewed, with most values concentrated between approximately 1.0 and 1.2 price units per 100g. The highest frequency, exceeding 1,500 occurrences, is observed around the 1.2 price mark. There are fewer products priced below 1.0, and very few exceed 1.5. This indicates that Voila’s prices are predominantly clustered around 1.2 per 100g, with some outliers on the lower and higher ends.

\newpage

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: fig-metro
#| fig-cap: Distribution of the Number of Becel Products Sold Across Different Prices from Metro

Metro_data <- subset(data, vendor == "Metro")


plot_Metro <- ggplot(Metro_data, aes(x = price_numeric)) +
  geom_histogram(binwidth = 0.1, fill = "blue", color = "black", alpha = 0.7) +
  labs(title = "Metro",
       x = "Price/100g",
       y = "Number of Products Bought") +
  theme_minimal()


print(plot_Metro)
```
@fig-metro represents the distribution of Metro’s price per 100g. The horizontal axis shows the price per 100g, while the vertical axis indicates the frequency of occurrences. The distribution is slightly right-skewed, with the majority of prices clustered between approximately 1.0 and 1.3 price units per 100g. The highest frequency, around 900 occurrences, is observed near the 1.2 price mark. A smaller number of items are priced below 1.0, and there are few outliers above 1.5. This suggests that Metro’s prices predominantly fall within the 1.0–1.3 range, with a peak at 1.2.

\newpage

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: fig-nofrills
#| fig-cap: Distribution of the Number of Becel Products Sold Across Different Prices from No Frills

NoFrills_data <- subset(data, vendor == "NoFrills")


plot_NoFrills <- ggplot(NoFrills_data, aes(x = price_numeric)) +
  geom_histogram(binwidth = 0.1, fill = "blue", color = "black", alpha = 0.7) +
  labs(title = "No Frills",
       x = "Price/100g",
       y = "Number of Products Bought") +
  theme_minimal()


print(plot_NoFrills)
```

@fig-nofrills illustrates the distribution of No Frills’ price per 100g. The horizontal axis represents the price per 100g, and the vertical axis indicates the frequency of occurrences. The distribution is bimodal, with significant peaks around 0.9 and 1.2 price units per 100g. The highest frequency, exceeding 1,000 occurrences, is observed at the 1.2 price mark. A smaller but notable peak is visible around 0.9, with fewer items priced below this range or above 1.5. This suggests that No Frills’ prices are primarily concentrated at two levels, with the higher peak being more dominant.


# Discussion {#sec-discussion}

The analysis of grocery product pricing data from Voila, Metro, and No Frills reveals a consistent pricing range for most products, particularly between $1.10 and $1.30 per 100g. This range suggests that these vendors are pricing their products competitively to appeal to cost-sensitive consumers in the Toronto market. The standardized approach, converting all product prices to a per-100g basis, provides a clear picture of pricing patterns across these stores, allowing for straightforward comparisons and a deeper understanding of each vendor's pricing strategy.

Focusing specifically on Becel products, the analysis found no direct correlation between the vendors in terms of pricing for this brand. This indicates that while the vendors may position themselves within a similar price range for other products, Becel pricing does not exhibit uniformity across retailers. Instead, each vendor appears to set Becel prices independently, potentially reflecting different sourcing agreements or sales strategies. However, the general consistency in the pricing range suggests a shared approach to targeting typical customer expectations, particularly for items with established price points.

These findings highlight that while there may not be a strict alignment in pricing specific brands, such as Becel, there is a broader trend of pricing standardization that caters to a competitive and customer-focused market. Future research could expand upon this by examining other brands or analyzing data over multiple time points to capture trends and fluctuations that single-point analysis cannot reveal. Such an approach would provide a more comprehensive picture of vendor strategies and how they adjust prices in response to market dynamics.

## A Note on Correlation vs. Causation 

As this analysis looks to examine the relationships between vendors in terms of pricing with Becel products, it focuses on whether there is a correlation between vendors and their pricing strategies. Causal conclusions cannot be made in the present analysis by reason that a careful experimental design was not employed to investigate whether one vendor's pricing is the result of another vendor's pricing decisions. To this, a future step may be to design an experiment that looks at how a change in one vendor's price for their Becel products impacts the subsequent prices decisions made by other vendors. 

## Missing Data

The possibility of having missing data within the original dataset provided by Project Hammer was briefly mentioned in @sec-data. This analysis narrows its focus to consider only three vendors as well as one specific brand, and for this reason, the analysis data that was used in the present project did not have any missing values. However, it is still important to consider the possibility of missing values within the original dataset as it is important to have an understanding of the source in which the data within this analysis comes from. 

## Sources of Bias

The data collected through screen-scraping is potentially subject to selection bias. Since the scraping process may not capture every product from each vendor, there could be unintentional omissions that skew the dataset. Additionally, focusing solely on three specific vendors—Voila, Metro, and No Frills—limits the generalizability of the findings. This selection bias may impact the representativeness of the pricing patterns observed and should be considered when interpreting the results.

## Limitations
This analysis is constrained by the observational nature of the dataset and the use of data from a single time point. Seasonal variations, temporary promotions, and price fluctuations that could affect customer purchasing decisions are not captured in this snapshot, limiting the conclusions to this specific timeframe. Moreover, while standardizing prices to 100g units allows for comparability, it may overlook vendor-specific pricing incentives such as multi-unit discounts, packaging variations, or bulk pricing, which can influence overall customer perception of value.

\newpage

# References